A Fast and Tight Heuristic for A* in Road Networks

### 2021-10-15 paper/ch_potentials.tex
move bidir switching exp into main body

In this section, we discuss a very specific setting arises naturally from A* heuristics.
Here, the target node is known in the selection step but the source nodes are only queried one after another.
We denote this problem as the \emph{incremental many-to-one} problem.
The first step is the \emph{target selection} where the target node is provided.
Then, an arbitrary number of source nodes $s_i$ are given one after another and for each source the distance to the target has to be computed.

@@

The Bucket-CH selection phase still explores the shared parts of the search space of the targets several times.
This can also be improved with the following modified CH backward search:
All target nodes are inserted into the queue at once.
The priority queue is ordered by importance instead of distance.
Distance labels are buckets instead if scalar values.
The buckets of each target node are initialized with a single entry for the respective node with distance zero.
When relaxing an edge, the bucket of the head nodes is merged with the bucket of the tail node with each distance increased by the weight of the edge.
By keeping the buckets sorted, this merging can be efficiently implemented with a coordinated linear sweep over both buckets.
We evaluate our Lazy RPHAST many-to-many algorithm with this improved Bucket-CH selection in Section~\ref{todo}.

@@

% TODO alternating directions

@@

% A* is a flexible algorithm.
% Many extended problems can therefore be solved using the CH-Potentials framework.
We start by establishing a common formal framework for the use of CH-Potentials.
Then, we exemplary describe some extended routing problems and how to apply CH-Potentials to them.

@@

@@

\begin{table}
\centering
\caption{
Performance of bidirectional and unidirectional A* on OSM Ger with different query weights.
The symmetric variant uses the improved pruning, the average variant does not.
All variants use all low degree optimizations.
}\label{tab:bidir}
\input{table/bidir.tex}
\end{table}
% TODO remove fourth scenario

In Table~\ref{tab:bidir} we investigate the effectiveness of bidirectional search for A* depending on the query weights.
Interestingly, only the zero heuristic and ALT consistently achieve speed-ups through bidirectional search on all query weights.
When the query weights are equal, CH-Potentials will already only traverse the shortest path with unidirectional search.
In this case it is clear that bidirectional search will only introduce unnecessary overhead.
When query weights are scaled up uniformly, bidirectional search achieves some search space reduction but it is not enough to significantly reduce running times due to the overhead of running a second search.
This changes significantly when the query weight increases are applied non-uniformly as in the third scenario.
Here only weights with a speed less than 80\,kph were scaled up.
This touches only the beginning and end of most shortest paths between randomly chosen nodes.
The middle part of the shortest paths will typically use faster edges like highways.
In this case bidirectional (C)CH-Potentials are a factor of two faster than the unidirectional variant.
The reason for this is that the search space of a unidirectional search expands greatly while exploring the end of the path to the target where the reduced weights are bad.
In contrast, the bidirectional searches meet in the middle of the shortest path where the reduced weights are close to zero and thus avoid this expansion.
This is also the reason why ALT behaves like this for all query weights.
By construction, the ALT heuristic has better reduced weights for edges which are on many shortest paths like highways.
This makes bidirectional search so critical for ALTs performance.
We conclude that having a potential as tight as CH-Potentials usually makes bidirectional search unnecessary.
Bidirectional search for A* only pays of when the reduced weights exhibit certain very specific patterns.

@@

\section{Extended Experimental Results on Bidrectional A*}

>>>>

In this section, we discuss a very specific setting which naturally arises from A* heuristics.
Here, the target node is known in the selection step but the source nodes are only queried one after another.
We denote this problem as the \emph{incremental many-to-one} problem.
The first step is the \emph{target selection} where the target node is provided.
Then, an arbitrary number of source nodes $s_i$ are given one after another and for each source the distance to the target has to be computed.

@@

The Bucket-CH selection phase still explores the shared parts of the search space of the targets several times.
This can also be improved with the following modified CH backward search:
All target nodes are inserted into the queue at once.
The priority queue is ordered by importance instead of distance.
Distance labels are buckets instead if scalar values.
The buckets of each target node are initialized with a single entry for the respective node with distance zero.
When relaxing an edge, the bucket of the head nodes is merged with the bucket of the tail node with each distance increased by the weight of the edge.
By keeping the buckets sorted, this merging can be efficiently implemented with a coordinated linear sweep over both buckets.
We evaluate our Lazy RPHAST many-to-many algorithm with this improved Bucket-CH selection in Section~\ref{sec:exp_lazy_rphast}.

@@

For both average and symmetric bidirectional A*, we always alternate between settling a node form the forward and the backward search.
In our experiments, we also evaluate choosing the direction by the smaller minimal queue key.
Even though this appears to better fit together with the stopping criterion for the symmetric approach, our experiments show that it is never beneficial in terms of running time (see Table~\ref{tab:bidir_switching}).

@@

% A* is a flexible algorithm.
Many extended problems can be solved using CH-Potentials.
We start by establishing a common formal framework for the use of CH-Potentials.
Then, we exemplary describe some extended routing problems and how to apply CH-Potentials to them.

@@

\subsection{Lazy RPHAST}\label{sec:exp_lazy_rphast}

@@

Table~\ref{tab:bidir_switching} shows the performance of bidirectional A* with different strategies to decide whether to advance the forward or the backward search next.
The results clearly show that alternating the directions is always superior.
Selecting the direction by minimal queue key may lead to huge imbalances in the progress of the searches.
This causes the searches to meet later and the total search space to grow significantly.

\begin{table}
\centering
\caption{
Performance of bidirectional and unidirectional A* on OSM Ger with different query weights.
The symmetric variant uses the improved pruning, the average variant does not.
All variants use all low degree optimizations.
}\label{tab:bidir}
\input{table/bidir.tex}
\end{table}

In Table~\ref{tab:bidir} we investigate the effectiveness of bidirectional search compared to unidirectional search depending on the query weights.
Interestingly, only the zero heuristic and ALT consistently achieve speed-ups through bidirectional search on all query weights.
When the query weights are equal, CH-Potentials will already only traverse the shortest path with unidirectional search.
In this case it is clear that bidirectional search will only introduce unnecessary overhead.
When query weights are scaled up uniformly, bidirectional search achieves some search space reduction but it is not enough to significantly reduce running times due to the overhead of running a second search.
This changes significantly when the query weight increases are applied non-uniformly as in the third scenario.
Here only weights with a speed less than 80\,kph were scaled up.
This touches only the beginning and end of most shortest paths between randomly chosen nodes.
The middle part of the shortest paths will typically use faster edges like highways.
In this case bidirectional (C)CH-Potentials are a factor of two faster than the unidirectional variant.
The reason for this is that the search space of a unidirectional search expands greatly while exploring the end of the path to the target where the reduced weights are bad.
In contrast, the bidirectional searches meet in the middle of the shortest path where the reduced weights are close to zero and thus avoid this expansion.
This is also the reason why ALT behaves like this for all query weights.
By construction, the ALT heuristic has better reduced weights for edges which are on many shortest paths like highways.
This makes bidirectional search so critical for ALTs performance.
We conclude that having a potential as tight as CH-Potentials usually makes bidirectional search unnecessary.
Bidirectional search for A* only pays of when the reduced weights exhibit certain very specific patterns.

@@
